The Weakly Nonlinear Saturation of the Magnetorotational Instability

### 2014-09-24 paper/mri_paper.tex

The MRI evolves via a period of linear growth, followed by nonlinear saturation. The nonlinear evolution of the MRI is a complex process, and an understanding of the mechanism by which the MRI saturates remains elusive. Most previous works on MRI saturation have studied the problem in simulation. Indeed, only a few works have tackled MRI saturation analytically. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. Another notable analytic work is that of \citeauthor{Pessah:2010ic} (\citeyear{Pessah:2010ic}), who presents an analysis of MRI saturation via secondary instabilities. This work suggests that MRI saturation can be caused by parasite instabilities growing off of the primary MRI modes, in particular by Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime.

@@

We develop this investigation with the goal of providing a rigorous analytical framework that will be used to make predictions for laboratory MRI experiments. Observing MRI in a laboratory setting would open a new avenue for studying MRI dynamics and saturation. In recent years much experimental progress has been made, although no experiment has yet yielded a confirmed detection. Of most immediate interest to this work is the Princeton Plasma Physics Laboratory (PPPL) MRI experiment, which has a design roughly analogous to the set-up investigated here (see \S\ref{setup} for further details). The Madison plasma dynamo experiment is a multi-purpose plasma device which may be a vehicle for MRI studies in the near future (\citei{Cooper:2013to}). The helical MRI, an instability related to the standard MRI but easier to excite in the laboratory, is the focus of the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE) in Dresden. PROMISE has already detected helical MRI in the laboratory (e.g. \citei{Stefani:2006iv}), and we discuss possible applications of our framework to the helical MRI in \S\ref{discussion}. A spherical Couette experiment claimed to have detected MRI in 2004 (\citeauthor{Sisan:2004ig}), but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}; \citei{Gissinger:2011td}).

The focus of this work is the weakly nonlinear regime of the MRI. This is where the initial linear growth phase of the instability begins to give way to nonlinear saturation. Weakly nonlinear analysis is the perturbation analysis of a system that is just unstable to its most unstable mode. The system in its marginal state, when the most unstable mode neither grows nor decays, is tuned slightly into instability. This excites a band of wavemodes around the most unstable mode, which interact nonlinearly. In order to track the evolution of the system on both rapid and slow scales, we conduct a formal multiscale analysis.

@@

\section{Set-Up}

@@

Our domain is designed to be relevant to the design of the PPPL MRI experiment (see \S\ref{discussion} for further discussion). We use periodic vertical boundary conditions. On the radial boundaries of our domain --  the inner and outer cylinders -- we apply no-slip, perfectly conducting boundary conditions. The choice of boundary conditions dictates the functional bases on which our system is spectrally solved. This is explained further in \S\ref{wna}. A diagram of our set-up is shown in Figure \ref{setup}.

@@

We perform a weakly nonlinear analysis following the lead of \citeauthor{Umurhan:2007hs}, and use much of the same notation, with some changes for added simplicity.

@@

\beq\label{momentum}
\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = 
\eeq

$\, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, + \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right)$

@@

\beq
\reye \equiv \frac{\Omega_0 r_0^2 \delta^2}{\nu},
\eeq

@@

\beq
\reym \equiv \frac{\Omega_0 r_0^2 \delta^2}{\eta}.
\eeq

@@

where $\alpha$ is a slowly-varying amplitude equation and $c.c.$ denotes the complex conjugate. The $x$ dependence is contained in $\mathbb{V}_{11} = (\Psi_{11}, u_{11}, A_{11}, B_{11})^\mathbf{T}$, and must be solved for subject to the radial boundary conditions. The periodic vertical boundary conditions allow us to posit the $z$ dependence, where $k_c$ is the value of the vertical wavenumber at marginality. As noted by \citeauthor{Umurhan:2007hs}, there exists a spatially constant neutral mode solution to the $B_y$ equation, with $\mathbb{U}_{11} = (0, 0, 0, 1)^\mathbf{T}$. The slowly-varying amplitude $\beta$ encodes the slow evolution of this mode.

@@

\begin{figure*}[h!]
\centering
\includegraphics[scale=0.45]{allorders_Bfield.eps}
\caption{First order (left) and second order (right) magnetic field perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. The colorbar represents azimuthal velocity.}\label{allorders_Bfield}
\end{figure*}

% perturbation expansion
The state vector is expanded a perturbation series in orders of $\epsilon$,

@@

The multiscale expansion of the operators and the perturbation series expansion of the fluid quantities combine to make the expansion of the nonlinear vector $\mathbf{N}$ expansive, and we relegate the expansion to Appendix \ref{app:nonlinear}. $\mathbf{N}$ becomes

@@

@@

The practical advantage of our ansatz construction (Equation \ref{V1_ansatz}) is clear: the separable x-dependence means that the radial boundary conditions are solved in only one dimension. Thus our analytical framework is able to side-step many of the resolution issues faced by multidimensional simulations. We are able to resolve even small-scale structure in the boundary layers of our domain, because we need only resolve it in one dimension. We solve the radial component of each equation using the open source pseudospectral code Dedalus (Burns et al., in prep). We compute the radial components on a grid of Chebyshev polynomials, as is appropriate for bounded one-dimensional domains (e.g. \citei{Boyd:2001aa}). The nonuniform spacing of the Chebyshev grid allows us to resolve the boundary structures very well on 256-point grid. See Figure \ref{N20fig} for an example of the boundary-layer resolution we achieve in a nonlinear term.

@@

where, again, $\mathbb{V}^\dagger$ contains the radial contribution. The elimination of this inner product must also be enforced on the terms in Equation \ref{ordere3} that are resonant with $e^{i k_c z}$, yielding

@@

but $\widetilde{c}$ is equal to zero and drops out of the equation. This is the well-known Ginzburg-Landau equation, and is with only slight modification the same amplitude equation derived in \citeauthor{Umurhan:2007hs}.

@@

but both $k$ and $p$ are equal to zero, so this amplitude equation decouples from Equation \ref{ampeq1}, and reveals itself to be a simple diffusion equation,

@@

\section{Parameter Space}

>>>>

The MRI evolves via a period of linear growth, followed by nonlinear saturation. The nonlinear evolution of the MRI is a complex process, and an understanding of the mechanism by which the MRI saturates remains elusive. Most previous works on MRI saturation have studied the problem in simulation. Indeed, only a few works have tackled MRI saturation analytically. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. Another notable analytic work is that of \citeauthor{Pessah:2010ic} (\citeyear{Pessah:2010ic}), who presents an analysis of MRI saturation via secondary instabilities. This work suggests that MRI saturation can be caused by parasitic instabilities growing off of the primary MRI modes, in particular by Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime.

@@

We develop this investigation with the goal of providing a rigorous analytical framework that will be used to make predictions for laboratory MRI experiments. Observing MRI in a laboratory setting would open a new avenue for studying MRI dynamics and saturation. In recent years much experimental progress has been made, although no experiment has yet yielded a confirmed detection. Of most immediate interest to this work is the Princeton Plasma Physics Laboratory (PPPL) MRI experiment, which has a design roughly analogous to the set-up investigated here (see \S\ref{sec:setup} for further details). The Madison plasma dynamo experiment is a multi-purpose plasma device which may be a vehicle for MRI studies in the near future (\citei{Cooper:2013to}). The helical MRI, an instability related to the standard MRI but easier to excite in the laboratory, is the focus of the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE) in Dresden. PROMISE has already detected helical MRI in the laboratory (e.g. \citei{Stefani:2006iv}), and we discuss possible applications of our framework to the helical MRI in \S\ref{discussion}. A spherical Couette experiment claimed to have detected MRI in 2004 (\citeauthor{Sisan:2004ig}), but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}; \citei{Gissinger:2011td}).

The focus of this work is the weakly nonlinear regime of the MRI. This is where the initial linear growth phase of the instability begins to give way to nonlinear saturation. Weakly nonlinear analysis is the perturbation analysis of a system that is just unstable to its most unstable mode. The system in its marginal state, when the most unstable mode neither grows nor decays, is tuned slightly into instability. This excites a band of wavemodes around the most unstable mode, which interact nonlinearly. In order to track the evolution of the system on both rapid and slow scales, we conduct a formal multiscale analysis. We develop and solve this framework in \S\ref{wna}, conduct a full parameter space study in \S\ref{sec:pspace}, and explore the implications of our findings in \S\ref{discussion}.

@@

\section{Set-Up}
\label{sec:setup}

@@

Our domain is designed to be relevant to the apparatus of the PPPL MRI experiment (see \S\ref{discussion} for further discussion). We use periodic vertical boundary conditions. On the radial boundaries of our domain --  the inner and outer cylinders -- we apply no-slip, perfectly conducting boundary conditions. The choice of boundary conditions dictates the functional bases on which our system is spectrally solved. This is explained further in \S\ref{wna}. A diagram of our set-up is shown in Figure \ref{setup}.

@@

We perform a weakly nonlinear analysis following the lead of \citeauthor{Umurhan:2007hs}, and use much of the same notation, with some changes for added clarity.

@@

\beq\label{momentum}
\begin{split}
& \partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = \, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, \\
& + \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right), \\
\end{split}
\eeq

@@

\beq
\reye \equiv \frac{\Omega_0 L^2}{\nu},
\eeq

@@

\beq
\reym \equiv \frac{\Omega_0 L^2}{\eta}.
\eeq

@@

where $\alpha$ is a slowly-varying amplitude equation and $c.c.$ denotes the complex conjugate. The $x$ dependence is contained in $\mathbb{V}_{11} = (\Psi_{11}, u_{11}, A_{11}, B_{11})^\mathbf{T}$, and must be solved subject to the radial boundary conditions. The periodic vertical boundary conditions allow us to posit the $z$ dependence, where $k_c$ is the value of the vertical wavenumber at marginality. As noted by \citeauthor{Umurhan:2007hs}, there exists a spatially constant neutral mode solution to the $B_y$ equation, with $\mathbb{U}_{11} = (0, 0, 0, 1)^\mathbf{T}$. The slowly-varying amplitude $\beta$ encodes the slow evolution of this mode.

@@

\begin{figure*}[h!]
\centering
\includegraphics[scale=0.45]{allorders_Bfield.eps}
\caption{First order (left) and second order (right) magnetic field perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. The colorbar represents azimuthal magnetic field strength.}\label{allorders_Bfield}
\end{figure*}

% perturbation expansion
The state vector is expanded in a perturbation series in orders of $\epsilon$,

@@

The multiscale expansion of the operators and the perturbation series expansion of the fluid quantities combine to make the expansion of the nonlinear vector $\mathbf{N}$ formidable, and we relegate the expansion to Appendix \ref{app:nonlinear}. $\mathbf{N}$ becomes

@@

In the above, we find that $\mathbf{V_2}$ takes the form

\beq
\begin{split}
\mathbf{V_2} \, & = \, \alpha^2 e^{i2Qz} \mathbb{V}_{22}(x) \, + \, \left|\alpha\right|^2 \mathbb{V}_{20}(x) \, \\
&- \, \partial_Z \alpha \mathbb{V}_{21}(x) e^{iQz} \,- \, \partial_Z \beta \mathbb{U}_{20}(x) + c.c.
\end{split}
\eeq

@@

The practical advantage of our ansatz construction (Equation \ref{V1_ansatz}) is clear: the separable x-dependence means that the radial boundary conditions are solved in only one dimension. Thus our analytical framework is able to side-step many of the resolution issues faced by multidimensional simulations. We are able to resolve even small-scale structure in the boundary layers of our domain, because we need only resolve it in one dimension. We solve the radial component of each equation using the open source pseudospectral code Dedalus (Burns et al., in prep). We compute the radial components on a grid of Chebyshev polynomials, as is appropriate for bounded one-dimensional domains (e.g. \citei{Boyd:2001aa}). The nonuniform spacing of the Chebyshev grid allows us to resolve the boundary structures well on a 256-point grid. See Figure \ref{N20fig} for an example of the boundary-layer resolution we achieve in a nonlinear term.

@@

where, again, $\mathbb{V}^\dagger$ contains the radial contribution. The elimination of this inner product must also be enforced on the terms in Equation \ref{ordere3} that are resonant with $e^{i k_c z}$, which are

\beq
\begin{split}
& \mathcal{D} \mathbb{V}_{11} e^{iQz}\partial_T \alpha \, + \, \alpha \left| \alpha^2 \right| \mathbf{N}_{31} e^{iQz} \, + \, \alpha \partial_Z \beta \widetilde{\mathbf{N}}_{31} e^{iQz} \, =  \\
& \, - \partial_Z^2 \alpha \widetilde{\mathcal{L}}_1 \mathbb{V}_{21} e^{iQz} \, + \, \partial_Z^2 \alpha \widetilde{\mathcal{L}}_2 \mathbb{V}_{11} e^{iQz} \, - \, \partial_Z \alpha \mathbb{V}_{11}\mathcal{X}e^{iQz} \, \\
& + \, iQ^3 \alpha \mathcal{L}_3 \mathbb{V}_{11} e^{iQz}.
\end{split}
\eeq

Taking the inner product yields the amplitude equation

@@

where ${}^*$ denotes a complex conjugate, but $\widetilde{c}$ is equal to zero and drops out of the equation. This is the well-known Ginzburg-Landau equation, and is with only slight modification the same amplitude equation derived in \citeauthor{Umurhan:2007hs}

@@

but both $k$ and $p$ are equal to zero. Thus this amplitude equation decouples from Equation \ref{ampeq1}, and reveals itself to be a simple diffusion equation,

@@

\section{Parameter Space}\label{sec:pspace}
